OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-05 paper/paper.Rmd
Start new sentences on new lines

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
in closed reference mode across any dataset/database combination. OptiFit was
able to fit **X** more query sequences to reference OTUs created with the
Greengenes and SILVA databases than with RDP. vsearch was able to fit **W%**
more query sequences to the Greengenes reference than OptiFit in closed
reference mode. In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and vsearch by **B%**.

@@

We developed a new algorithm for fitting sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering. OptiFit makes the
iterative method employed by OptiClust available for tasks where
reference-based clustering is required. We have shown that OTU quality is
similar between OptiClust and OptiFit in open reference mode, regardless of
strategy employed. open reference OptiFit does perform slower than OptiClust
due to the additional _de novo_ clustering step, so users may prefer OptiClust
for tasks that do not require reference OTUs.

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which has only about 16,000 V4
16S sequences compared to about 170,000 in SILVA and Greengenes. We recommend
users who require an independent reference database opt for large databases with
good coverage of microbial diversity. Since OptiClust performs faster than open
reference OptiFit and creates higher quality OTUs than closed reference OptiFit
with the database strategy, we recommend using OptiClust rather than fitting to
a database whenever stable OTUs are not required for the study at hand.

The mothur algorithms produced higher quality OTUs than vsearch in open
reference, closed reference, or _de novo_ modes. However, vsearch was able to
fit more sequence into OTUs than OptiFit in closed reference mode. While both
mothur and vsearch use a distance or similarity threshold for determining how to
assign sequences to OTUs, vsearch is more permissive than mothur. Mothur's
OptiFit and OptiClust use all of the sequences to define an OTU, requiring that
all pairs of sequences (including reference and query sequences) in an OTU are
within the distance threshold without penalizing the MCC. In contrast, vsearch
only requires each query sequence to be similar to the single sequence that
seeded the OTU. In this way, vsearch sacrifices OTU quality in order to allow
more sequences to fit to OTUs. Users who require closed reference clustering to
the Greengenes database may prefer to use vsearch if they wish to maximize the
fraction of sequences that can be fit at the cost of OTU quality. However,
mothur's OptiClust or OptiFit are recommended for _de novo_ or open reference
clustering to produce OTUs of the highest possible quality.

@@

- **TODO: big picture concluding paragraph**

>>>>

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
fitting to RDP.
Only up to **Z%** of query sequences were fit to reference OTUs in closed
reference mode across any dataset/database combination.
OptiFit was able to fit **X** more query sequences to reference OTUs created
with the Greengenes and SILVA databases than with RDP.
Vsearch was able to fit **W%** more query sequences to the Greengenes reference
than OptiFit in closed reference mode.
In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and vsearch by **B%**.

@@

We developed a new algorithm for fitting sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering.
OptiFit makes the iterative method employed by OptiClust available for tasks
where reference-based clustering is required.
We have shown that OTU quality is similar between OptiClust and OptiFit in open
reference mode, regardless of strategy employed.
Open reference OptiFit performs slower than OptiClust due to the additional _de
novo_ clustering step, so users may prefer OptiClust for tasks that do not
require reference OTUs.

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which has only about 16,000 V4
16S sequences compared to about 170,000 in SILVA and Greengenes (**TODO: insert
R code to generate these numbers**).
We recommend that users who require an independent reference database opt for
large databases with good coverage of microbial diversity.
Since OptiClust performs faster than open reference OptiFit and creates higher
quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever stable OTUs
are not required for the study at hand.

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than vsearch in open reference, closed reference, or _de novo_ modes.
However, vsearch was able to fit more sequence into OTUs than OptiFit in closed
reference mode.
While both mothur and vsearch use a distance or similarity threshold for
determining how to assign sequences to OTUs, vsearch is more permissive than
mothur.
Mothur's OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, vsearch only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, vsearch sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use vsearch if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

@@

**TODO: big picture concluding paragraph.**
We have developed a new reference-based clustering algorithm that allows users
to produce high quality OTUs when stability is a requirement for the task at hand.
